Understanding School Shootings: Analyzing Trends, Demographics, and Geographic Factors Using Bayesian Methods

### 2024-12-04 paper/paper.qmd
update the format and the table in data section
update the format and the table in data section

---
title: ""
subtitle: ""
author: 
  - Ruizi Liu
thanks: "Code and data are available at: [https://github.com/RIRI0527/school_shooting_analysis.git](https://github.com/RIRI0527/school_shooting_analysis.git)."
date: today
date-format: long
abstract: "School shootings remain a critical issue in the United States, with far-reaching social and psychological consequences. This study investigates the factors influencing the occurrence of school shootings, focusing on temporal trends, school type, demographic composition, and geographic context. Using a Bayesian logistic regression model, this analysis incorporates both observed and simulated data to explore the relationship between these predictors and the likelihood of a shooting. The results reveal significant associations between the predictors and school shootings. Temporal trends indicate an alarming increase in incidents over time, while public schools and urban locales exhibit higher risks compared to private schools and rural areas. Demographic composition also plays a role, with schools characterized by specific racial majorities showing distinct patterns in the likelihood of shootings. This report emphasizes the need for targeted, data-driven interventions that consider institutional and geographic variations in risk. By leveraging probabilistic modeling, this study provides actionable insights for policymakers aiming to mitigate the occurrence of school shootings and improve the safety of educational environments across the United States. "
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

We used the`R`programming language [@citeR], the `dplyr` package[@R_dplyr], the `tidyr` package [@R_tidyr], the `maps` package [@R_maps], the `ggplot2` package [@R_ggplot2], and the`scale` package [@R_scales] to clean the data.
In doing so, R code was adapted from [@tellingstories].

@@

## Outcome variables
The dataset reveals a notable disparity in the distribution of school types. As shown in @fig-school-type, public schools account for the vast majority of cases, while private schools are significantly underrepresented. This imbalance reflects the broader landscape of education in the United States, where public schools serve a larger student population. Moreover, this distribution is critical for our analysis, as it highlights the need to account for institutional differences when examining factors contributing to school shootings.

@@

# Create a bar plot
ggplot(school_type_counts, aes(x = reorder(school_type, -Count), y = Count)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(title = "Distribution of School Types",
       x = "School Type",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@fig-ulocale illustrates the distribution of urban locales where school shootings occurred. Large cities and large suburbs are overrepresented compared to smaller towns and remote rural areas. This trend aligns with population density and resource disparities, suggesting that urban schools may face heightened risks due to greater exposure to social and economic stressors. Understanding the geographic distribution of incidents provides essential context for identifying at-risk locations and tailoring preventive measures accordingly.

@@

# Assuming your dataset is loaded into a data frame called 'data'
# Count occurrences of each category in 'ulocale_desc'
ulocale_counts <- as.data.frame(table(shooting_data$ulocale_desc))

# Rename the columns for clarity
colnames(ulocale_counts) <- c("ulocale_desc", "Count")

# Create a bar plot
ggplot(ulocale_counts, aes(x = reorder(ulocale_desc, -Count), y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  labs(x = "Urban Locale Description",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

@@

## Appendix

### A Additional Data Details

@@

### B Model Details

#### B.1 Posterior Predictive Check

@@

#### B.2 Diagnostics

@@

# Visualize Rhat values
bayesplot::mcmc_rhat(rhat_vals) +
  theme_classic() +
  ggtitle("Rhat Diagnostic Plot")
```
#### B.3 Comparison of Priors and Posteriors

@@

### C Survey and Sampling Methodology

#### C.1 Population, Frame, and Sample
- **Population:** All K-12 schools in the United States.
- **Frame:** Schools from the Washington Post dataset, combined with synthetic data for schools without shootings.
- **Sample:** Stratified by school type, urban locale description, and dominant demographic group.

#### C.2 Sampling and Recruitment
- Synthetic data were generated within observed variable ranges to represent schools where shootings did not occur.
- Observed data were supplemented by demographic, geographic, and enrollment metrics.

#### C.3 Questionnaire and Validation
- **Survey Focus:** School security measures, demographic composition, geographic details, and historical incidents.
- **Validation:** Cross-referenced data with administrative records to ensure accuracy.

### D Simulation Details

To simulate the occurrence of school shootings, synthetic data were generated with random variability:
- Geographic coordinates were adjusted within observed ranges.
- Demographics and enrollment were assigned proportionally based on observed distributions.

@fig-occurrence-map visualizes the geographic distribution of shooting occurrences, highlighting both observed and synthetic data points.

@@

### E Implications of Results

2. **Institutional and Demographic Context:** Public schools and urban locales were associated with higher risks, as shown in @fig-credible-intervals.

@@

These findings provide actionable insights for policymakers aiming to mitigate risks associated with school shootings.

>>>>

---
title: "Understanding School Shootings: Analyzing Trends, Demographics, and Geographic Factors Using Bayesian Methods"
subtitle: "A Data-Driven Investigation into the Predictors of School Shootings in the United States (1999–2024)"
author: 
  - Ruizi Liu
thanks: "Code and data are available at: [https://github.com/RIRI0527/school_shooting_analysis.git](https://github.com/RIRI0527/school_shooting_analysis.git)."
date: today
date-format: long
abstract: "School shootings remain a critical issue in the United States, with far-reaching social and psychological consequences. This study investigates the factors influencing the occurrence of school shootings, focusing on temporal trends, school type, demographic composition, and geographic context. Using a Bayesian logistic regression model, this analysis incorporates both observed and simulated data to explore the relationship between these predictors and the likelihood of a shooting. The results reveal significant associations between the predictors and school shootings. Temporal trends indicate an alarming increase in incidents over time, while public schools and urban locales exhibit higher risks compared to private schools and rural areas. Demographic composition also plays a role, with schools characterized by specific racial majorities showing distinct patterns in the likelihood of shootings. This report emphasizes the need for targeted, data-driven interventions that consider institutional and geographic variations in risk. By leveraging probabilistic modeling, this study provides actionable insights for policymakers aiming to mitigate the occurrence of school shootings and improve the safety of educational environments across the United States. "
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

We used the`R`programming language [@citeR], the `here` package [],
the `dplyr` package [@R_dplyr], the `tidyr` package [@R_tidyr], the `maps` package [@R_maps], the `ggplot2` package [@R_ggplot2], and the`scale` package [@R_scales] to clean the data. We used the `tidyverse` package [@tidyverse], the `maps` package [@maps], the `ggplot2` package [@R_ggplot2], the `knitr` package [@knitr], the `kableExtra` package [@kableExtra] to plot the dataset and the visualization.

In doing so, R code was adapted from [@tellingstories].

@@

```{r}
#| label: tlb-data-overview
#| tib-cap: The specific columns in the dataset
#| echo: false
#| warning: false

# Subset the dataset for demonstration (first 10 rows and selected columns)
summary_data <- data[1:10, c("year", "state", "school_type", "top_1_races", "ulocale_desc", "data_source", "school_shooting")]

# Create the kable table
kable(summary_data, 
      col.names = c("Year", "State", "School Type", "Top 1 Race", "Ulocale", "Data Source", "School Shooting"),
      caption = "Summary of School Shooting Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "left")
```

## Outcome variables

@@

# Create a bar plot
ggplot(school_type_counts, aes(x = reorder(school_type, -Count), y = Count)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(
       x = "School Type",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The dataset reveals a notable disparity in the distribution of school types. As shown in @fig-school-type, public schools account for the vast majority of cases, while private schools are significantly underrepresented. This imbalance reflects the broader landscape of education in the United States, where public schools serve a larger student population. Moreover, this distribution is critical for our analysis, as it highlights the need to account for institutional differences when examining factors contributing to school shootings.

@@

filtered_data <- data %>%
  filter(data_source == "raw" & !is.na(ulocale_desc))

ggplot(filtered_data, aes(x = ulocale_desc, fill = ulocale_desc)) +
  geom_bar() +
  geom_text(stat = "count", 
            aes(label = paste0(round(..count../sum(..count..) * 100, 1), "%")), 
            vjust = -0.2, color = "black", size = 3.5) +
  scale_fill_viridis_d(option = "D") + # Use a visually appealing viridis palette
  labs(
    x = "Urban Locale Description",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) + # Set base font size for overall consistency
  theme(
    legend.position = "none", # Remove legend
    plot.title = element_text(face = "bold", hjust = 0.5), # Center-align title
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels for readability
  )
```

@fig-ulocale descriptions for schools with raw data highlights key geographic trends. As visualized in the bar chart, Suburb: Large and City: Large locales account for the majority of school shootings, reflecting the higher density of schools and populations in these areas. Smaller categories, such as Rural: Remote and Town: Fringe, are underrepresented, suggesting that sparsely populated regions experience fewer incidents. This pattern underscores the role of urbanicity in shaping exposure to risk factors like population density, socioeconomic conditions, and resource allocation. The findings emphasize the need for geographically tailored interventions, with a focus on urban and suburban areas that bear a disproportionate burden of school shootings. These insights can guide policymakers in targeting resources and preventive measures effectively.

@@

We use the `brms` package [@brms] and the `bayesplot` package [@bayesplot] to fit the model.

@@

# Appendix

## A Additional Data Details

@@

## B Model Details

### B.1 Posterior Predictive Check

@@

### B.2 Diagnostics

@@

# Visualize Rhat values
bayesplot::mcmc_rhat(rhat_vals) +
  theme_classic() +
  ggtitle("Rhat Diagnostic Plot")
```
### B.3 Comparison of Priors and Posteriors

@@

## C Survey and Sampling Methodology

### C.1 Population, Frame, and Sample
- **Population:** All K-12 schools in the United States.
- **Frame:** Schools from the Washington Post dataset, combined with synthetic data for schools without shootings.
- **Sample:** Stratified by school type, urban locale description, and dominant demographic group.

### C.2 Sampling and Recruitment
- Synthetic data were generated within observed variable ranges to represent schools where shootings did not occur.
- Observed data were supplemented by demographic, geographic, and enrollment metrics.

### C.3 Questionnaire and Validation
- **Survey Focus:** School security measures, demographic composition, geographic details, and historical incidents.
- **Validation:** Cross-referenced data with administrative records to ensure accuracy.

## D Simulation Details

@@

```
To simulate the occurrence of school shootings, synthetic data were generated with random variability:
- Geographic coordinates were adjusted within observed ranges.
- Demographics and enrollment were assigned proportionally based on observed distributions.

@fig-occurrence-map visualizes the geographic distribution of shooting occurrences, highlighting both observed and synthetic data points.
## E Implications of Results

@@

2. **Institutional and Demographic Context:** Public schools and urban locales were associated with higher risks, as shown in @fig-credible-intervals.
These findings provide actionable insights for policymakers aiming to mitigate risks associated with school shootings.
